---
output:
  pdf_document: default
  html_document: default
---
```{r}
library(HardyWeinberg)
library(data.table)
genotype_data = fread("./TSIChr22v4.raw")
df <- data.frame(genotype_data)
df <- df[, (7:ncol(df))]; head(df)
```

# 1. How many variants are there in this database? What percentage of the data is missing?
```{r}
n_variants = ncol(df)
cat("Number of variants in the dataset:", n_variants)

# Convert non-0, 1, or 2 values to NA
df[!sapply(df, function(x) x %in% c(0, 1, 2))] <- NA

n_missing = sum(is.na(df))
total_data_points = ncol(df) * nrow(df)
perc_missing = (n_missing / total_data_points)*100
cat("\nPercentages missing data:", perc_missing)
```
# 2. Calculate the percentage of monomorphic variants. Exclude all monomorphics from the database for all posterior computations of the practical. How many variants do remain in your database?
```{r}
monomorphic <- sapply(df, function(x) length(unique(na.omit(x))) == 1)
cat("Percentage of monomorphic variants:", (sum(monomorphic)/n_variants)*100, "%\n")
df_mono <- df[, !monomorphic]
cat(ncol(df_mono), "variants remain in the database")
```

# 3. Extract polymorphism rs587756191 T from the data, and determine its genotype counts. Apply a chi-square test for Hardy-Weinberg equilibrium, with and without continuity correction. Also try an exact test, and a permutation test. You can use thee functions HWChisq, HWExact and HWPerm for this purpose. Do you think this variant is in equilibrium?

```{r}
specific_variant <- df_mono[,"rs587756191_T"]
genotype_counts <- c(
  AA=sum(specific_variant==0),
  AB=sum(specific_variant==1),
  BB=sum(specific_variant==2)
)

# Chi-square test without continuity correction
hw_chisq <- HWChisq(genotype_counts, cc=0)

# Chi-square test with continuity correction
hw_chisq_corrected <- HWChisq(genotype_counts)

# Exact test for Hardy-Weinberg Equilibrium
hw_exact <- HWExact(genotype_counts)

# Permutation test for Hardy-Weinberg Equilibrium
hw_perm <- HWPerm(genotype_counts)

```
