---
title: "BSG-MDS practical 5 Statistical Genetics"
author: 
  - Pyry Satama
  - Max de Visser
subtitle: "05/12/2023, submission deadline 12/12/2023"
output:
  pdf_document: default
  html_document: default
---

# 1. Perform the alleles test for this data set. Provide the p-value and the odds ratio and comment on the results.
```{r}
X <- matrix(c(112, 278, 150, 206, 348, 150),byrow=TRUE,ncol=3)
colnames(X) <- c("AA", "Aa", "aa")
rownames(X) <- c("Cases", "Controls")
X
Y <- cbind(2*X[,1]+X[,2],2*X[,3]+X[,2]);
colnames(Y) <- c("A","a");Y

chisq_test <- chisq.test(Y,correct=FALSE)
cat("\nP-value:", chisq_test$p.value)
odds_ratio_test <- fisher.test(Y)
cat("\nOdds ratio:", odds_ratio_test$estimate)
```
The small p-value is highly significant indicating strong evidence against the null hypothesis of no association between the allele and Alzheimer's disease.

The odds ratio less than 1, suggests that the presence of the allele under consideration is associated with a reduced odds of having Alzheimer's disease

# 2. Test for association using a codominant, a dominant and a recessive model. Provide the p-values for all the tests and comment on the results.
```{r}
# Codominant model
chisq_test_codominant <- chisq.test(X)
p_value_codominant_chi <- chisq_test_codominant$p.value
p_value_codominant_fish <- fisher.test(X)$p.value

# Dominant model
Y_dominant <- cbind(X[,1], X[,2]+X[,3])
colnames(Y_dominant) <- c("AA", "Aa or aa")
rownames(Y_dominant) <- c("Cases", "Control")
p_value_dominant_chi <- chisq.test(Y_dominant)$p.value
p_value_dominant_fish <- fisher.test(Y_dominant)$p.value

# Recessive model
Y_recessive <- cbind(X[,1]+X[,2],X[,3])
colnames(Y_recessive) <- c("AA or Aa", "aa")
rownames(Y_recessive) <- c("Cases", "Control")
p_value_recessive_chi <- chisq.test(Y_recessive)$p.value
p_value_recessive_fish <- fisher.test(Y_recessive)$p.value

cat("P-value of Chi-squared test for codominant model:", p_value_codominant_chi, "\n")
cat("P-value of Fisher's test for codominant model:", p_value_codominant_fish, "\n\n")

cat("P-value of Chi-squared test for dominant model:", p_value_dominant_chi, "\n")
cat("P-value of Fisher's test for dominant model:", p_value_dominant_fish, "\n\n")
cat("P-value of Chi-squared testfor recessive model:", p_value_recessive_chi, "\n")
cat("P-value of Fisher's test for recessive model:", p_value_recessive_fish, "\n")
```
P-values for codominant model suggest very strong association between Alzheimer's and the genotypes when considering each genotype separately.

P-values of dominant model also suggest an strong association under the dominant model, where individuals with at least one A allele (AA or Aa) are compared to those with none (aa). This suggests that carrying at least one A allele could be linked to an increased or decreased risk of Alzheimer's.

P-values of recessive model are higher that for other models but still suggest a statistically significant association under the recessive model. The relatively higher p-value in this model compared to the others suggests that the recessive model may be a less strong fit for the data, but there is still evidence to suggest that the aa genotype is associated with Alzheimer's.

# 3. Plot the risk of disease as a function of the number of m alleles. Comment on the results. Which model seems most appropriate?
```{r}
MM_cases <- X["Cases", "AA"]
Mm_cases <- X["Cases", "Aa"]
mm_cases <- X["Cases", "aa"]

MM_controls <- X["Controls", "AA"]
Mm_controls <- X["Controls", "Aa"]
mm_controls <- X["Controls", "aa"]

total_MM <- MM_cases + MM_controls
total_Mm <- Mm_cases + Mm_controls
total_mm <- mm_cases + mm_controls

risk_MM <- MM_cases / total_MM
risk_Mm <- Mm_cases / total_Mm
risk_mm <- mm_cases / total_mm

m_alleles <- c(0, 1, 2)

risks <- c(risk_MM, risk_Mm, risk_mm)

plot(m_alleles, risks, type="b", xlab="Number of m alleles", ylab="Risk of Disease", 
     main="Risk of Disease vs. Number of m Alleles")
```
Based on the plot the risk increases consistently with the number of m alleles. Thus additive model seems most appropriate.