---
title: "BSG-MDS practical 5 Statistical Genetics"
author: 
  - Pyry Satama
  - Max de Visser
subtitle: "05/12/2023, submission deadline 12/12/2023"
output:
  pdf_document: default
  html_document: default
---

```{r}
library(data.table)
library(ggplot2)
library(gdsfmt)
library(SNPRelate)
library(dplyr)
```

```{r}
df <- fread("YRI6.raw")
df <- data.frame(df, header = TRUE)
genetic_info <- df[, (7:ncol(df))]
```

# 1. How many individuals and how many SNPs are there in the database? What percentage of the data is missing?
```{r}
num_individuals <- nrow(genetic_info)
num_SNPs <- ncol(genetic_info)
cat("Number of individuals:", num_individuals)
cat("\nNumber of SNPs:", num_SNPs)

n_missing <- sum(is.na(genetic_info))
total_data_points <- num_individuals * num_SNPs
perc_missing <- (n_missing / total_data_points)*100
cat("\nPercentages missing data:", perc_missing, "%")
```

# 2. Compute, for each pair of individuals (and report the first 5), the mean m of the number of alleles shared and the standard deviation s of the number of alleles shared.
```{r}
shared_alleles <- function(a, b) {
  # if a-b = 0 (no difference), then 2 is returned
  # if a-b = 2 (largest difference), then 0 is returned
  return(2-abs(a-b))
}

# To store results:
num_pairs <- choose(num_individuals, 2) # number of combinations
results <- data.frame(ind1 = character(num_pairs), ind2 = character(num_pairs), m = numeric(num_pairs), s = numeric(num_pairs))

iids <- df$IID

k <- 1
for(i in 1:(num_individuals-1)) {
  for(j in (i+1):num_individuals) {
    pairs_shared <- apply(genetic_info[c(i,j), ], 2, function(x) shared_alleles(x[1], x[2]))
    m = mean(pairs_shared)
    s = sd(pairs_shared)
    
    # df[i, ]$IID is the ith individuals ID
    # df[j, ]$IID is the jth individuals ID
    results[k, ] <- c(iids[i], iids[j], m, s)
    k <- k + 1
  }
}

head(results, 5)
```

# 3. Compute, for each pair of individuals (and report the first 5), the fraction of variants for which the individuals share 0 alleles (p0), and the fraction of variants for which the individuals share 2 alleles (p2). Check if m = 1 − p0 + p2 holds.
```{r}
results2 <- data.frame(ind1 = character(num_pairs), ind2 = character(num_pairs), p0 = numeric(num_pairs), p2 = numeric(num_pairs), m = numeric(num_pairs), check_m = logical(num_pairs)) # to store the results

k <- 1
for(i in 1:(num_individuals - 1)) {
  for(j in (i + 1):num_individuals) {
    pairs_shared <- apply(genetic_info[c(i, j), ], 2, function(x) shared_alleles(x[1], x[2]))
    
    # Calculate p0 and p2
    p0 <- mean(pairs_shared == 0)
    p2 <- mean(pairs_shared == 2)
    m <- mean(pairs_shared)
    
    # Check if m = 1 - p0 + p2
    check_m <- abs(m - (1 - p0 + p2)) < 0.0001 # account for small numerical errors
    
    
    # df[i, ]$IID is the ith individuals ID
    # df[j, ]$IID is the jth individuals ID
    results2[k, ] <- c(iids[i], iids[j], p0, p2, m, check_m)
    k <- k + 1
  }
}

head(results2, 5)
```

# 4.  Plot m against s and plot p0 against p2. Comment on the results.
```{r}
results_copy <- data.frame(results)[, c("m", "s")]
results_copy$m <- as.numeric(results_copy$m)
results_copy$s <- as.numeric(results_copy$s)
results_rounded <- round(results_copy, 8)

# Plot m against s (standard deviation)
ggplot(results_rounded, aes(x = m, y = s)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Mean vs. Standard Deviation of Shared Alleles",
       x = "Mean of Shared Alleles (m)",
       y = "Standard Deviation of Shared Alleles (s)")

results2_copy <- data.frame(results2)[, c("p0", "p2")]
results2_copy$p0 <- as.numeric(results2_copy$p0)
results2_copy$p2 <- as.numeric(results2_copy$p2)
results2_rounded <- round(results2_copy, 8)

# Plot p0 against p2
ggplot(results2_rounded, aes(x = p0, y = p2)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Fraction of 0 Alleles Shared vs. 2 Alleles Shared",
       x = "Fraction of 0 Alleles Shared (p0)",
       y = "Fraction of 2 Alleles Shared (p2)")
```
For the Mean vs. Standard deviation, we see two distance clusters appear indicating that the data reflects the presence of a subpopulation. Same is present for the 0/2 alleles sharing plot; we see roughly the same pattern with two clusters emerging. The cluster in the upper left corner is always at least sharing one allele (fraction of 0 shared alleles is 0), and a bit less than half of the time they are sharing two alleles indicating that this cluster contains similar genetic information. The cluster in the lower right is more often sharing 0 alleles (about 15% of the times), and they share 2 alleles less frequently then upper left cluster.

# 5. Plot m against s and use the pedigree information of the YRI06.raw file to label the data points in the scatterplot. Recall that column 3 and 4 from the YRI06.raw contain information about the family relationship of the participants. Create two labels: one for individuals that have a parent-offspring relationship and another one for unrelated individuals. Comment on the results.
```{r}
related <- function(ind1_ID, ind2_ID) {
  # Function to check if two individuals have a parent offspring relationship.
  ind1_info <- df[df$IID == ind1_ID, c("PAT", "MAT")]
  ind2_info <- df[df$IID == ind2_ID, c("PAT", "MAT")]
  if (ind1_info$PAT == ind2_ID | ind1_info$MAT == ind2_ID | ind2_info$PAT == ind1_ID | ind2_info$MAT == ind1_ID) {
    return (TRUE) # the pair of individuals are related
 } else {
    return (FALSE) # they are not related
  }
}

results_rounded$related <- apply(results[, c("ind1", "ind2")], 1, function(row) related(row[1], row[2]))

# Plot m against s (standard deviation)
ggplot(results_rounded, aes(x = m, y = s, color = related)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Mean vs. Standard Deviation of Shared Alleles",
       x = "Mean of Shared Alleles (m)",
       y = "Standard Deviation of Shared Alleles (s)")
```
In the plot, we observe that people who are related have a consistently higher mean of shared alleles between individuals and also a consistently lower standard deviation hereof, which supports the idea of closely related individuals share more genetic information. On the other hand, unrelated individuals are, relative to the related individuals, a bit more scattered both in terms of the mean and the standard deviation. The means of the unrelated individuals is lower and the standard deviation is higher indicating higher variance in the number of shared alleles between unrelated individuals and thus less genetic information shared between individuals.

# 6. Use the package SNPRelate to estimate the IBD probabilities, and plot the probabilities of sharing 0 and 1 IBD alleles (k0 and k1) for all pairs of individuals. Use the pedigree information of the YRI06.raw file to label the data points in the scatterplot (same as before, one colour for parent-offspring relationship and another colour for unrelated individuals).

```{r}
# create GDS file: 
gds_data <- snpgdsCreateGeno("gds_file.gds", 
                             genmat=as.matrix(genetic_info), 
                             sample.id = df$IID, 
                             snp.id=colnames(genetic_info),
                             snpfirstdim=FALSE)
genofile <- snpgdsOpen("gds_file.gds") # load GDS file

ibd <- snpgdsIBDMLE(genofile, maf=0.05, num.thread = 2)
```
```{r}
k0_data <- data.frame(ibd$k0) # probability of sharing 0 IBD alleles
k1_data <- data.frame(ibd$k1) # probability of sharing 1 IBD allele

# To store results
ibd_results <- data.frame(ind1 = character(num_pairs), ind2 = character(num_pairs), k0 = numeric(num_pairs), k1 = numeric(num_pairs))

# Loop over unique pair of values in the upper triangular part of both matrices
# and store the probabilities in a dataframe
k0_rows <- nrow(k0_data)
k0_cols <- ncol(k0_data)
k <- 0
for (i in 1:(k0_rows - 1)) {
  for (j in (i + 1):k0_cols) {
    ind1_ <- iids[i]
    ind2_ <- iids[j]
    k0_prob <- k0_data[i, j]
    k1_prob <- k1_data[i, j]
    
    ibd_results[k, ] <- c(ind1_, ind2_, k0_prob, k1_prob)
    k <- k + 1
  }
}

# created related variable for the dataframe
ibd_results <- ibd_results[-nrow(ibd_results), ] # remove last empty row 
ibd_results$related <- apply(ibd_results[, c("ind1", "ind2")], 1, function(row) related(row[1], row[2]))

# round k0 and k1
precision <- 2
ibd_results$k0 <- as.numeric(ibd_results$k0, digits = precision)
ibd_results$k0 <- round(ibd_results$k0, digits = precision)
ibd_results$k1 <- as.numeric(ibd_results$k1, digits = precision)
ibd_results$k1 <- round(ibd_results$k1, digits = precision)

ggplot(ibd_results, aes(x = k0, y = k1, color = related)) +
  geom_point() +
  theme_minimal() +
  labs(title = "k0 vs. k1",
       x = "k0",
       y = "k1")
```

# 7. Do you think the family relationships between all individuals were correctly specified?

Given the plot with k0 against k1, we observe that if individuals are related through a parent-off spring relation, then k1 is always 1 and k0 is always 0 (all blue points are clustered in the upper left corner). This reflects the fact that, for related individuals, at least 1 IBD allele is inherited from one of the parents - which is generally true. The obtained results for parent offspring related individuals also align with our expected values for k0 and k1 which are 0 and 1, respectively.

In the case of unrelated individuals (red points) we see slightly more scatter of points, but in general all points cluster in the lower right corner which indicates that for unrelated individuals in general k0=1, meaning that in almost all cases of comparisons, there is a high probability of sharing no alleles inherited from a recent common ancestor. Additionally, in all comparisons between unrelated individuals, there is a low probability of sharing one allele inherited from a recent common ancestor. The obtained results for unrelated individuals mostly align with our expected values for k0 and k1 which are 1 and 0, respectively. There are few outliers which have higher values of k1 (red points with k1 > ~0.01), indicating some errors. However, most points are clusters in the lower right corner.

As a result, we believe that almost all family relationships between individuals were correctly specified (with the exception of the few outliers perhaps.)
