---
title: "BSG-MDS practical 5 Statistical Genetics"
author: 
  - Pyry Satama
  - Max de Visser
subtitle: "05/12/2023, submission deadline 12/12/2023"
output:
  pdf_document: default
  html_document: default
---

```{r}
library(data.table)
library(ggplot2)
```

```{r}
df <- fread("YRI6.raw")
df <- data.frame(df, header = TRUE)
genetic_info <- df[, (7:ncol(df))]
```

# 1. How many individuals and how many SNPs are there in the database? What percentage of the data is missing?
```{r}
num_individuals <- nrow(genetic_info)
num_SNPs <- ncol(genetic_info)
cat("Number of individuals:", num_individuals)
cat("\nNumber of SNPs:", num_SNPs)

n_missing <- sum(is.na(genetic_info))
total_data_points <- num_individuals * num_SNPs
perc_missing <- (n_missing / total_data_points)*100
cat("\nPercentages missing data:", perc_missing, "%")
```

# 2. Compute, for each pair of individuals (and report the first 5), the mean m of the number of alleles shared and the standard deviation s of the number of alleles shared.
```{r}
shared_alleles <- function(a, b) {
  return(2-abs(a-b))
}

results <- data.frame(Pair = character(), m = numeric(), s = numeric())

for(i in 1:(num_individuals-1)) {
  for(j in (i+1):num_individuals) {
    pairs_shared <- apply(genetic_info[c(i,j), ], 2, function(x) shared_alleles(x[1], x[2]))
    m = mean(pairs_shared)
    s = sd(pairs_shared)
    
    results <- rbind(results, data.frame(Pair = paste("Ind", i, "- Ind", j), m = m, s = s))
  }
}

head(results, 5)
```

# 3. Compute, for each pair of individuals (and report the first 5), the fraction of variants for which the individuals share 0 alleles (p0), and the fraction of variants for which the individuals share 2 alleles (p2). Check if m = 1 − p0 + p2 holds.
```{r}
results2 <- data.frame(Pair = character(), p0 = numeric(), p2 = numeric(), m = numeric(), check_m = logical())

for(i in 1:(num_individuals - 1)) {
  for(j in (i + 1):num_individuals) {
    pairs_shared <- apply(genetic_info[c(i, j), ], 2, function(x) shared_alleles(x[1], x[2]))
    
    # Calculate p0 and p2
    p0 <- mean(pairs_shared == 0)
    p2 <- mean(pairs_shared == 2)
    m <- mean(pairs_shared)
    
    # Check if m = 1 - p0 + p2
    check_m <- abs(m - (1 - p0 + p2)) < 0.0001

    results2 <- rbind(results2, data.frame(Pair = paste("Ind", i, "- Ind", j), p0 = p0, p2 = p2, m = m, check_m = check_m))
  }
}

head(results2, 5)
```

# 4.  Plot m against s and plot p0 against p2. Comment on the results.
```{r}
# Plot m against s (standard deviation)
ggplot(results, aes(x = m, y = s)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Mean vs. Standard Deviation of Shared Alleles",
       x = "Mean of Shared Alleles (m)",
       y = "Standard Deviation of Shared Alleles (s)")

# Plot p0 against p2
ggplot(results2, aes(x = p0, y = p2)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Fraction of 0 Alleles Shared vs. 2 Alleles Shared",
       x = "Fraction of 0 Alleles Shared (p0)",
       y = "Fraction of 2 Alleles Shared (p2)")
```